Practical Optional Types for Clojure

### 2015-07-06 overview.tex
Change end of section 2

The \clj{hi} multimethod dispatches on its arguments via the identity function.

@@

HMap types model the most common usages of keyword maps.
The inferred type for \clj{breakfast} holds two kinds of information---the known entries \clj{:en}
and \clj{:fr} have string values (in \clj{:mandatory}) and that 
every other key is absent (in \clj{:complete?}).

HMap's default to being partially specified. The HMap shorthand annotating
\clj{lunch}---below---forgets 
information about absent keys, only providing \clj{:mandatory}
information.

\begin{minted}{clojure}
(ann lunch '{:en Str :fr Str})
(def lunch {:en "muffin" :fr "baguette"})

(:en lunch)    ; :- Str
;=> "muffin"
(:fr lunch)    ; :- Str
;=> "baguette"
; Unknown lookups are now less accurate
(:bocce lunch) ; :- Any
;=> nil
\end{minted}

\paragraph{HMaps in Practice} The next example is extracted from a production system at CircleCI,
a company with a large production Typed Clojure system
(section~\ref{sec:casestudy} presents a case study).

@@

(ann enc-keypair [RawKeyPair -> EncKeyPair])
(defn enc-keypair "Encrypt an unencrypted keypair"
  [{pk :private-key :as kp}] ; original map is kp
  (assoc 
    ; remove unencrypted private key
    (dissoc kp :private-key)
    ; add encrypted private key
    :enc-private-key (encrypt pk)))
\end{minted}
%\begin{exmp}
%\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
%\label{example:circleci}
%\end{exmp}
Fully specified HMap's are used to
statically enforce the private key is not accidentally left in a supposedly
encrypted keypair,

\begin{minted}{clojure}
(ann enc-keypair-bad [RawKeyPair -> EncKeyPair])
(defn enc-keypair-bad
  [{pk :private-key :as kp}]
  (assoc kp :enc-private-key (encrypt pk)))
\end{minted}

@@

Partially specified HMap's with a common dispatch key
combine naturally with ad-hoc unions.

\begin{minted}{clojure}
(defalias Order   ; define type abbreviation
  "A meal order, tracking dessert quantities."
  (U '{:Meal ':lunch ; keyword singleton type
       :desserts Num}
     '{:Meal ':dinner :desserts Num}
     '{:Meal ':combo :meal1 Order :meal2 Order}))

(ann desserts [Order -> Int])
(defmulti desserts "Number of desserts per order."
          :Meal)  ; dispatch on :Meal entry
; map destructuring backwards, d is :desserts entry
(defmethod desserts :lunch [{d :desserts}] d)
(defmethod desserts :dinner [{d :desserts}] d)
(defmethod desserts :combo [{m1 :meal1 m2 :meal2]
  (+ (desserts m1) (desserts m2)))

(cutlery {:Meal :combo 
          :meal1 {:Meal :lunch :desserts 1}
          :meal2 {:Meal :dinner :desserts 2}})
;=> 3
\end{minted}

@@

\paragraph{HMap dispatch} The flexibility of \clj{isa?} is key to the generality of multimethods. 
In \egref{example:incmap} we
dispatch on the \clj{:op} key 
of our HMap AST \clj{Expr}.
Since keywords are functions that look themselves up in their argument, we simply
use \clj{:op} as the dispatch function.

\begin{exmp}
\inputminted[firstline=5,lastline=18]{clojure}{code/demo/src/demo/eg5.clj}
\label{example:incmap}
\end{exmp}

The function \clj{inc-leaf} is like \egref{example:decmap} except the nodes are incremented.
The reasoning is similar, except we only consider one branch (the current method) by
locally considering the current \emph{dispatch value} and reasoning about how it relates
to the \emph{dispatch function}.
For example, 
in the \clj{:do} method we learn the \clj{:op} key is a \clj{:do}, which
narrows our argument type to the \clj{:do} Expr, and similarly for the \clj{:if}
and \clj{:const} methods.

\subsection{Final example}

\egref{example:final}
combines everything we will cover for the rest of the paper:
multimethod dispatch, reflection resolution via type hints, Java method
and constructor calls, conditional and exceptional flow reasoning,
and HMaps.

\begin{figure}
\begin{exmp}
\inputminted[firstline=6,lastline=23]{clojure}{code/demo/src/demo/eg7.clj}
\label{example:multidispatch}
\end{exmp}
\begin{exmp}
\inputminted[firstline=6,lastline=20]{clojure}{code/demo/src/demo/eg8.clj}
\label{example:final}
\end{exmp}
\caption{Multimethod Examples}
\end{figure}

We dispatch on \clj{:p} to distinguish the two cases of \clj{FSM}---for example on \clj{:F}
we know the \clj{:file} is a file.
The body of the first method uses type hints to resolve reflection
and conditional control flow to prove null-pointer exceptions are impossible.
The second method is similar except it uses exceptional control flow.

>>>>

The \clj{hi} multimethod dispatches on its argument via the identity function.

@@

HMap types model the most common usages of keyword maps.
The inferred type for \clj{breakfast} holds two kinds of information---the known entries \clj{:en}
and \clj{:fr} have string values (expressed as a \clj{:mandatory} entry) and that 
every other key is absent (by \clj{:complete?} being \clj{true}).

HMap's default to being partially specified---\clj{:complete?} defaults to \clj{false}. The HMap shorthand 
\clj{'{:en Str :fr Str}}
forgets 
information about absent keys, only providing \clj{:mandatory}
information.

\begin{minted}{clojure}
(ann lunch '{:en Str :fr Str})
(def lunch {:en "muffin" :fr "baguette"})
(:en lunch)    ; :- Str
;=> "muffin"
(:fr lunch)    ; :- Str
;=> "baguette"
; Unknown lookups are now less accurate
(:bocce lunch) ; :- Any
;=> nil
\end{minted}

\paragraph{HMaps in practice} The next example is extracted from a production system at CircleCI,
a company with a large production Typed Clojure system
(section~\ref{sec:casestudy} presents a case study).

@@

(ann enc-keypair [RawKeyPair -> EncKeyPair])
(defn enc-keypair "Encrypt an unencrypted keypair"
  [{pk :private-key :as kp}] ; original map is kp
  (assoc 
    ; remove unencrypted private key
    (dissoc kp :private-key)
    ; add encrypted private key
    :enc-private-key (encrypt pk)))
\end{minted}
%\begin{exmp}
%\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
%\label{example:circleci}
%\end{exmp}
Fully specified HMap's
statically enforce the private key is not accidentally left in a supposedly
encrypted keypair.

\begin{minted}{clojure}
(ann enc-keypair-bad [RawKeyPair -> EncKeyPair])
(defn enc-keypair-bad
  [{pk :private-key :as kp}]
  (assoc kp :enc-private-key (encrypt pk))
  ; Type Error:
  ;   Expected EncKeyPair, given 
  ;   (HMap :mandatory {:enc-private-key EncKey
  ;                     :private-key RawKey
  ;                     :public-key RawKey}
  ;         :complete? true)
  )
\end{minted}

The extra \clj{:private-key} entry disagrees with \clj{EncKeyPair}, so a type error
is given.

@@

\subsection{HMaps and multimethods, joined at the hip}

Since HMaps are the primary way of specifying the structure of data in Clojure,
and multimethods are the primary tool for dispatching on data, they are inevitably
linked.
There are infinite ways of both structuring and dispatching on data, so there is no
hope predicting them all, or even manually covering useful a subset.

Thankfully, occurrence typing is both extensible \emph{and} compositional.
By extending occurrence typing with
a handful of rules based on HMaps and other functions, 
we can automatically cover almost all the common cases---as well others
that compose simple rules in arbitrary ways.

Futhermore, no special attention is needed for multimethod dispatch---the primitive branching
mechanism is still the humble \clj{if} conditional. Only a small number of rules are needed
to encode the \clj{isa?}-based dispatch, themselves made of small, simple pieces.
In practice, this means all conditional-based control flow understood by occurrence typing
also extends to the context of multimethod dispatch, and vice-versa.

We first demonstrate a very common, simple dispatch style,
then move on to deeper structural dispatching where occurrence typing's
compositionality shines.

\paragraph{HMaps and unions} Partially specified HMap's with a common dispatch key
combine naturally with ad-hoc unions.
An \clj{Order} is one of three kinds of HMaps.

\begin{minted}{clojure}
(defalias Order   ; define type abbreviation
  "A meal order, tracking dessert quantities."
  (U '{:Meal ':lunch ; keyword singleton type
       :desserts Int}
     '{:Meal ':dinner :desserts Int}
     '{:Meal ':combo :meal1 Order :meal2 Order}))
\end{minted}

The \clj{:Meal} entry is common to each HMap, always mapped to a known keyword singleton
type.
It's natural to dispatch on the \clj{class} of an instance---it's similarly
natural to dispatch on a known entry like \clj{:Meal}.

\begin{minted}{clojure}
(ann desserts [Order -> Int])
(defmulti desserts "Number of desserts per order."
          :Meal)  ; dispatch on :Meal entry
; map destructuring backwards, d is :desserts entry
(defmethod desserts :lunch [{d :desserts}] d)
(defmethod desserts :dinner [{d :desserts}] d)
(defmethod desserts :combo [{m1 :meal1 m2 :meal2}]
  (+ (desserts m1) (desserts m2)))

(desserts {:Meal :combo 
           :meal1 {:Meal :lunch :desserts 1}
           :meal2 {:Meal :dinner :desserts 2}})
;=> 3
\end{minted}

The \clj{:combo} method is verified to only structurally recur
on \clj{Order}s. This is achieved because we learn the argument
must be at least of type
\clj{'{:Meal :combo}}
since
\clj{(isa? (:Meal o) :combo)}
must be true. This eliminates possibility of \clj{:lunch} and \clj{:dinner}
orders, deducing
\begin{minted}{clojure}
'{:Meal ':combo :meal1 Order :meal2 Order}
\end{minted}
as the argument type, which can safely structurally recur.

\paragraph{Arbitrary dispatch}
An equally valid dispatch mechanism for \clj{desserts}
would be on the \clj{class} of the \clj{:desserts} key.
We have already seen dispatch on \clj{class} and on keywords
in isolation---occurrence typing automatically understands
control flow that combines its simple building blocks.

\begin{minted}{clojure}
(ann desserts' [Order -> Int])
(defmulti desserts' 
          (fn [o :- Order] (class (:desserts o))))
(defmethod desserts' Long [{d :desserts}] d)
(defmethod desserts' nil [{m1 :meal1 m2 :meal2}]
  (+ (desserts' m1) (desserts' m2)))

(desserts' {:Meal :combo 
            :meal1 {:Meal :lunch :desserts 1}
            :meal2 {:Meal :dinner :desserts 2}})
;=> 3
\end{minted}

In the \clj{Long} method, Typed Clojure learns that
its argument is at least of type \clj{'{:desserts Long}}---since
\begin{minted}{clojure}
(isa? (class (:desserts o)) Long)
\end{minted}
must be true.

Combined with the knowledge that it is also an
\clj{Order}, we can eliminate the possibility of
a \clj{:combo} meal, and infer the type
\begin{minted}{clojure}
(U '{:Meal ':lunch :desserts Int}
   '{:Meal ':dinner :desserts Int})
\end{minted}
Looking up \clj{:desserts} on this type returns an \clj{Int}.

In the \clj{nil} method, we know
\begin{minted}{clojure}
(isa? (class (:desserts o)) nil)
\end{minted}
is true---which implies \clj{(class (:desserts o))} is \clj{nil}.

Since lookups on missing keys return \clj{nil}, there are
now two possibilities:
\begin{itemize}
  \item either \clj{o} has a \clj{:desserts} entry to \clj{nil}, or
  \item \clj{o} is missing a \clj{:desserts} entry.
\end{itemize}
In a type, we learn \clj{o} is at least
\begin{minted}{clojure}
(U '{:desserts nil}
   ; :absent-keys, a set of known absent entries
   (HMap :absent-keys #{:desserts}))
\end{minted}

This is enough to infer \clj{o} is a \clj{:combo} meal, thus verifying both
structural recursions.

@@

%\paragraph{HMap dispatch} The flexibility of \clj{isa?} is key to the generality of multimethods. 
%In \egref{example:incmap} we
%dispatch on the \clj{:op} key 
%of our HMap AST \clj{Expr}.
%Since keywords are functions that look themselves up in their argument, we simply
%use \clj{:op} as the dispatch function.
%
%\begin{exmp}
%\inputminted[firstline=5,lastline=18]{clojure}{code/demo/src/demo/eg5.clj}
%\label{example:incmap}
%\end{exmp}
%
%The function \clj{inc-leaf} is like \egref{example:decmap} except the nodes are incremented.
%The reasoning is similar, except we only consider one branch (the current method) by
%locally considering the current \emph{dispatch value} and reasoning about how it relates
%to the \emph{dispatch function}.
%For example, 
%in the \clj{:do} method we learn the \clj{:op} key is a \clj{:do}, which
%narrows our argument type to the \clj{:do} Expr, and similarly for the \clj{:if}
%and \clj{:const} methods.
%
%
%\subsection{Final example}
%
%\egref{example:final}
%combines everything we will cover for the rest of the paper:
%multimethod dispatch, reflection resolution via type hints, Java method
%and constructor calls, conditional and exceptional flow reasoning,
%and HMaps. 
%
%
%\begin{figure}
%\begin{exmp}
%\inputminted[firstline=6,lastline=23]{clojure}{code/demo/src/demo/eg7.clj}
%\label{example:multidispatch}
%\end{exmp}
%\begin{exmp}
%\inputminted[firstline=6,lastline=20]{clojure}{code/demo/src/demo/eg8.clj}
%\label{example:final}
%\end{exmp}
%\caption{Multimethod Examples}
%\end{figure}
%
%We dispatch on \clj{:p} to distinguish the two cases of \clj{FSM}---for example on \clj{:F}
%we know the \clj{:file} is a file.
%The body of the first method uses type hints to resolve reflection
%and conditional control flow to prove null-pointer exceptions are impossible.
%The second method is similar except it uses exceptional control flow.
